---
title: "Problem Set 3. Due Thurs March 2 5pm"
author: "Etienne Danis"
date: "Feb 23, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

# Submission

1. Fork this repository to your own account
1. Make changes to the `README.Rmd` file (including the author field above).
1. Knit the file to HTML format and add it to your git repository (`git add`)
1. Submit a pull request with your Rmd and knitted HTML files.

# Overview 

You will examine a [time-course of mRNA expression][1] of barley powdery mildew fungus in immunocompromised plants (*Arabidopsis*). The fungus Blumeria graminis f. sp. hordei (called *Bgh* from now on)

Counts were generated using the [`maSigPro` software][3] and deposited in [NCBI GEO][2].

[1]: http://www.ncbi.nlm.nih.gov/pubmed/23175786
[2]: http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE43163  
[3]: http://bioinformatics.oxfordjournals.org/content/30/18/2598.full

Some important identifiers for the data set:

| Name | Description |
| ---- | ----------- |
| `A6` | *Bgh* isolate expressing other AVRA effectors |
| `K1` | *Bgh* isolate expressing the cognate AVRA1 effector for MLA1|
| `pps` | *Arabidopsis* plants |
| `B12` | *Arabidopsis* plants expressing MLA1-HA|

We will download tables from GEO containing counts of mRNA abundance from both the fungal sample.

# Raw data

First we need load the data.

```{r load_raw_data, message = FALSE}
library(tidyverse)

# bgh data
bgh_url <- 'http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE43163&format=file&file=GSE43163_CompleteCountTable_Bgh.txt.gz'

# 2 info and blank lines at top, skip them
raw_data <- read_tsv(bgh_url, skip = 2)

# the header for the first column is "NA", set it manually
names(raw_data)[1] <- 'gene.name'

raw_data <- as_data_frame(raw_data)
raw_data
```

# Problems


## Problem 1

Now that the raw data are loaded, your first task is to tidy the data with `tidyr`. The `cleaned_data` should look something like this:

```
# A tibble: 310,896 × 6
             gene.name plant fungus time.point   rep value
*                <chr> <chr>  <chr>      <chr> <chr> <int>
1             bgh04079   B12     A6       6hpi     1    13
2             bgh01634   B12     A6       6hpi     1    31
3  bghG000012000001001   B12     A6       6hpi     1   121
4  bghG000012000002001   B12     A6       6hpi     1     3
5             bgh00757   B12     A6       6hpi     1   253
6             bgh01273   B12     A6       6hpi     1    45
7             bgh01274   B12     A6       6hpi     1    32
8             bgh01277   B12     A6       6hpi     1     9
9             bgh06140   B12     A6       6hpi     1    47
10            bgh05774   B12     A6       6hpi     1    42
# ... with 310,886 more rows
```

The key steps are to `gather()` the data into key-value pairs and then `separate()` the information from the key column into new columns.

```{r problem_1}
# add tidying code here
library(tidyverse)
raw_data %>% gather(key,value,-gene.name) %>% separate(key, into = c('plant', 'fungus', 'time.point', 'rep'), sep = '_') -> cleaned_data
cleaned_data

# report the cleaned_data by just naming it, uncomment the following line:
# cleaned_data
```

## Problem 2

You need to translate what you learned on the command line into the R world.

Which plant has the highest expression of any gene in the 6hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "6hpi"' | sort -k6nr | head -n 1 | cut -f2
```

```{r}
# translate to dplyr code

filter(cleaned_data, time.point=="6hpi") %>% arrange(desc(value)) %>% select(plant) %>%  filter(row_number()==1)

```

Which plant / fungus pair has the highest expression in the 18hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "18hpi"' | sort -k6nr | head -n 1 | cut -f2,3
```

```{r}
# translate to dplyr code

filter(cleaned_data, time.point=="18hpi") %>% arrange(desc(value)) %>% select(plant, fungus) %>% filter(row_number()==1)

```

## Problem 3

Identify the top 3 most consistently differentially expressed genes between the earliest and latest time points for each combination of `plant` and `fungus` strains.

- "Differential expression"" is the difference between `value` (i.e., gene expression level) between time points.

- "Consistency" is the smallest variance in `value` between replicates.

### Strategy

1. Create a new table from the cleaned data by moving each `hpi` value to a new column name, with counts for each in the column (hint: use a `tidyr` verb). It is helpful to reformat the `hpi` values by converting from e.g. `6hpi` to `hpi.6`. You can use `mutate` to do this, i.e.:

```{r reformat_time, eval=FALSE}
library(stringr)
# Version_1
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), time.point = str_c('hpi.', time.point))-> cleaned_data_hpi
cleaned_data_hpi
# Version_2: cleaned_data %>% mutate(time.value = str_replace(time.point, 'hpi', ''), time.hpi = str_c('hpi.', time.value)) %>% select(-time.point, -time.value)-> cleaned_data_hpi
```

2. Create a new column containing the expression difference between the relevant time points.

```{r new_column}
library(stringr)
# Version_1
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), time.point = str_c('hpi.', time.point))-> cleaned_data_hpi
cleaned_data_hpi %>% spread(time.point,value) %>% mutate(expr.diff = abs(hpi.24 - hpi.6), expr.diff.var = var(expr.diff)) -> expr_data
expr_data
```

3. Calculate summary statistics (mean and variance) of the expression differences by grouping (hint) the `gene.name`, and `plant` columns.


```{r substraction}
# Using substraction to determine the differentially expressed genes
expr_data %>% group_by(plant, fungus, gene.name) %>% summarise(expr.diff.mean = mean(expr.diff), expr.diff.var=var(expr.diff)) -> expr_data_mean 
expr_data_mean 
```

4. Sort by these statistics and use the `dplyr` verb `slice` to pull the ones you want (i.e., the top 3). Note you will have to remove `gene.name` from the grouping so that sorting works.

```{r}
# Previous method used: I sorted first by expr.diff.mean and then by expr.diff.var. This gave me very small gene expression variation so I changed to sort/arrange by expr.diff.var first
#expr_data_mean %>% arrange(desc(expr.diff.mean)) %>%  arrange(expr.diff.var) %>% slice(1:3) %>% select(plant, fungus, gene.name, expr.diff.mean, expr.diff.var) %>% group_by(plant,fungus,gene.name) -> top3_genes
#top3_genes %>% tbl_df() %>% print (n=12)
```

```{r}
# Sort by expr.diff.mean last
expr_data_mean %>% arrange(expr.diff.var) %>% arrange(desc(expr.diff.mean)) %>%  slice(1:3) %>% select(plant, fungus, gene.name, expr.diff.mean, expr.diff.var) %>% group_by(plant,fungus,gene.name) -> top3_genes
top3_genes %>% tbl_df() %>% print (n=12)
```
## Problem 4

Now examine the above final data frame above and write a few sentences putting inline code in least 3 places. For example, There are `r nrow(mtcars)` rows of data in `mtcars`.


```{r, echo = FALSE}

filter(top3_genes, grepl("B12", plant)) %>%  filter(., grepl("A6", fungus)) %>%  ungroup() %>% select(expr.diff.mean) %>%  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_B12_A6

filter(top3_genes, grepl("pps", plant)) %>%  filter(., grepl("A6", fungus)) %>%  ungroup() %>% select(expr.diff.mean) %>%  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_pps_A6

filter(top3_genes, grepl("B12", plant)) %>%  filter(., grepl("K1", fungus)) %>%  ungroup() %>% select(expr.diff.mean) %>%  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_B12_K1

filter(top3_genes, grepl("pps", plant)) %>%  filter(., grepl("K1", fungus)) %>%  ungroup() %>% select(expr.diff.mean) %>%  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_pps_K1

#top3_genes_for_B12_A6
top3_genes %>% subset(., plant =="B12" & fungus =="A6", select=c(gene.name)) %>% t()-> top3_genes_for_B12_A6t
top3_genes %>% subset(., plant =="B12" & fungus =="K1", select=c(gene.name)) %>% t()-> top3_genes_for_B12_K1t
top3_genes %>% subset(., plant =="pps" & fungus =="A6", select=c(gene.name)) %>% t()-> top3_genes_for_pps_A6t
top3_genes %>% subset(., plant =="pps" & fungus =="K1", select=c(gene.name)) %>% t()-> top3_genes_for_pps_K1t

```

My tidy dataset has `r ncol(top3_genes)` columns and `r nrow(top3_genes)` rows. The name of the columns are `r colnames(top3_genes)`.

The highest difference in expression of a gene in the B12 plant expressing the immune receptor MLA1-HA treated with other effectors than the AVRA1 effector is: `r top_gene_B12_A6`.

The highest difference im expression of a gene in the pps plant (partially immunocompromised) treated with effectors different from the AVRA1 effector is: `r top_gene_pps_A6`.

The highest difference in expression of a gene in the B12 plant expressing the immune receptor MLA1-HA treated with the AVRA1 effector is: `r top_gene_B12_K1`.

The highest difference in expression of a gene in the pps plant (partially immunocompromised) treated with effectors different from the AVRA1 effector is: `r top_gene_pps_K1`.

The top 3 genes for the pair (B12 plant + A6 fungus) are: `r top3_genes_for_B12_A6t`.

The top 3 genes for the pair (B12 plant + K1 fungus) are: `r top3_genes_for_B12_K1t`.

The top 3 genes for the pair (B12 plant + A6 fungus) are: `r top3_genes_for_pps_A6t`.

The top 3 genes for the pair (B12 plant + K1 fungus) are: `r top3_genes_for_pps_K1t`.


## Problem 5

Plot the expression (`value`) by time (`hpi`) for the above genes. Format the plot as follows:

```{r preliminary_for_the_graphs}
# 3 top genes per pair plant/fungus, 4 pairs, triplicates, 4 time points = 3 x 4 x 3 x 4 = 144
# First, create for each pair plant/fungus, the list of top 3 genes differentially expressed and save the info in different files
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', '')) %>%  select(gene.name, plant, fungus, rep, time.point, value) %>% group_by(plant,fungus,gene.name)-> data_selected

top3_genes %>% subset(., plant =="B12" & fungus =="A6", select=c(plant, fungus, gene.name)) -> top3_genes_for_B12_A6
top3_genes %>% subset(., plant =="B12" & fungus =="K1", select=c(plant, fungus, gene.name)) -> top3_genes_for_B12_K1
top3_genes %>% subset(., plant =="pps" & fungus =="A6", select=c(plant, fungus, gene.name)) -> top3_genes_for_pps_A6
top3_genes %>% subset(., plant =="pps" & fungus =="K1", select=c(plant, fungus, gene.name)) -> top3_genes_for_pps_K1

# Collect the time points and replicate values for the top 3 genes for each pair plant/fungus
# For the B12 plant and the A6 fungus
filter(data_selected, gene.name %in% top3_genes_for_B12_A6$gene.name) -> data_filtered_by_gene_name_B12_A6
filter(data_filtered_by_gene_name_B12_A6, plant %in% top3_genes_for_B12_A6$plant) -> data_filtered_by_gene_name_and_plant_B12_A6
filter(data_filtered_by_gene_name_and_plant_B12_A6, fungus %in% top3_genes_for_B12_A6$fungus) -> data_for_B12_and_A6_pair

# For the B12 plant and the K1 fungus
filter(data_selected, gene.name %in% top3_genes_for_B12_K1$gene.name) -> data_filtered_by_gene_name_B12_K1
filter(data_filtered_by_gene_name_B12_K1, plant %in% top3_genes_for_B12_K1$plant) -> data_filtered_by_gene_name_and_plant_B12_K1
filter(data_filtered_by_gene_name_and_plant_B12_K1, fungus %in% top3_genes_for_B12_K1$fungus) -> data_for_B12_and_K1_pair

# For the pps plant and the A6 fungus
filter(data_selected, gene.name %in% top3_genes_for_pps_A6$gene.name) -> data_filtered_by_gene_name_pps_A6
filter(data_filtered_by_gene_name_pps_A6, plant %in% top3_genes_for_pps_A6$plant) -> data_filtered_by_gene_name_and_plant_pps_A6
filter(data_filtered_by_gene_name_and_plant_pps_A6, fungus %in% top3_genes_for_pps_A6$fungus) -> data_for_pps_and_A6_pair

# For the pps plant and the K1 fungus
filter(data_selected, gene.name %in% top3_genes_for_pps_K1$gene.name) -> data_filtered_by_gene_name_pps_K1
filter(data_filtered_by_gene_name_pps_K1, plant %in% top3_genes_for_pps_K1$plant) -> data_filtered_by_gene_name_and_plant_pps_K1
filter(data_filtered_by_gene_name_and_plant_pps_K1, fungus %in% top3_genes_for_pps_K1$fungus) -> data_for_pps_and_K1_pair

#Combine all the dataframes in 1 for the graphs
all_data_for_graphs <- rbind(data_for_B12_and_A6_pair, data_for_B12_and_K1_pair, data_for_pps_and_A6_pair, data_for_pps_and_K1_pair)

```

1. Plot each point.
```{r question_5_1}
library(ggplot2)
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value)) + geom_point() + xlab("time")

```
2. Connect the points with a smooth line 
```{r question_5_2}
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, group = gene.name)) + geom_point()  + xlab("time") + geom_smooth(method = 'lm', se = F)
```
3. Create a facet for each `plant` / `bgh` pair 
```{r question_5_3}
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, group = gene.name)) + geom_point() + xlab("time") + facet_wrap(~plant~fungus) + geom_smooth(method = 'lm', se = F)
```
4. Assign a different color to each gene 
```{r question_5_4}
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, color = gene.name, group = gene.name)) + geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + geom_smooth(method = 'lm', se = F)
```

5. Assign a different shape to each replicate 

```{r question_5_5}
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, color = gene.name, group = gene.name, shape=factor(rep))) + geom_point()  + xlab("time") + facet_wrap(~plant~fungus) +  scale_shape_manual(values=c(21,24,8)) 
```

6. Add a descriptive title to the plot 

```{r question_5_6}
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, color = gene.name, group = gene.name, shape=factor(rep))) + geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + scale_shape_manual(values=c(21,24,8)) + ggtitle('Gene expression profile of the top 3 differentially \n expressed genes per pair plant/fungus over the time course')
```

7. Use the "black & white" theme

```{r question_5_7}
library(cowplot)
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, color = gene.name, group = gene.name, shape=factor(rep))) + geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + scale_shape_manual(values=c(21,24,8)) + ggtitle('Gene expression profile of the top 3 differentially \n expressed genes per pair plant/fungus over the time course') + theme_bw() + theme_cowplot()
```

Extra credit: add error bars to the plot (use `geom_errorbar`).

```{r 5_8_extra_credit}
#expr_data %>% group_by(plant, fungus, gene.name) %>% summarise(expr.diff.mean = mean(expr.diff), expr.diff.var=var(expr.diff), standard.deviation = sd(expr.diff)) -> expr_data_mean_for_sd 
#expr_data_mean_for_sd %>% arrange(desc(expr.diff.mean)) %>%  arrange(expr.diff.var) %>% slice(1:3) %>% select(plant, fungus, gene.name, expr.diff.mean, expr.diff.var, standard.deviation) %>% group_by(plant,fungus,gene.name) -> top3_genes_for_sd
#top3_genes_for_sd
#all_data_for_graphs %>% group_by(gene.name, plant, fungus, time.point) %>% summarise(mean.data = mean(value), standard.deviation = sd(value), ymin = mean.data - standard.deviation, ymax = mean.data + standard.deviation) -> all_data_for_graphs_with_sd 
#all_data_for_graphs_with_sd 
#ggplot(all_data_for_graphs_with_sd , aes(x = factor(time.point), y = standard.deviation)) + geom_point()
#all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
#ggplot(all_data_for_graphs_with_sd, aes(x = factor(time.point), y = value, color = gene.name, group = gene.name, shape=factor(rep))) + geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + scale_shape_manual(values=c(21,24,8)) + ggtitle('Gene expression profile of the top 3 differentially \n expressed genes per pair plant/fungus over the time course') + theme_bw() + theme_cowplot() + geom_errobar(aes())
```
