---
title: "Problem Set 3. Due Thurs March 2 5pm"
author: "Etienne Danis"
date: "Feb 23, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

# Submission

1. Fork this repository to your own account
1. Make changes to the `README.Rmd` file (including the author field above).
1. Knit the file to HTML format and add it to your git repository (`git add`)
1. Submit a pull request with your Rmd and knitted HTML files.

# Overview 

You will examine a [time-course of mRNA expression][1] of barley powdery mildew fungus 
in immunocompromised plants (*Arabidopsis*). 
The fungus Blumeria graminis f. sp. hordei (called *Bgh* from now on)

Counts were generated using the [`maSigPro` software][3] and deposited in [NCBI GEO][2].

[1]: http://www.ncbi.nlm.nih.gov/pubmed/23175786
[2]: http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE43163  
[3]: http://bioinformatics.oxfordjournals.org/content/30/18/2598.full

Some important identifiers for the data set:

| Name | Description |
| ---- | ----------- |
| `A6` | *Bgh* isolate expressing other AVRA effectors |
| `K1` | *Bgh* isolate expressing the cognate AVRA1 effector for MLA1|
| `pps` | *Arabidopsis* plants |
| `B12` | *Arabidopsis* plants expressing MLA1-HA|

We will download tables from GEO containing counts of mRNA abundance
from both the fungal sample.

# Raw data

First we need load the data.

```{r load_raw_data, message = FALSE}
library(tidyverse)

# bgh data
bgh_url <- 
  'http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE43163&format=file&file=GSE43163_CompleteCountTable_Bgh.txt.gz'

# 2 info and blank lines at top, skip them
raw_data <- read_tsv(bgh_url, skip = 2)

# the header for the first column is "NA", set it manually
names(raw_data)[1] <- 'gene.name'

raw_data <- as_data_frame(raw_data)
raw_data
```

# Problems


## Problem 1

Now that the raw data are loaded, your first task is to tidy the data with `tidyr`. 
The `cleaned_data` should look something like this:

```
# A tibble: 310,896 × 6
             gene.name plant fungus time.point   rep value
*                <chr> <chr>  <chr>      <chr> <chr> <int>
1             bgh04079   B12     A6       6hpi     1    13
2             bgh01634   B12     A6       6hpi     1    31
3  bghG000012000001001   B12     A6       6hpi     1   121
4  bghG000012000002001   B12     A6       6hpi     1     3
5             bgh00757   B12     A6       6hpi     1   253
6             bgh01273   B12     A6       6hpi     1    45
7             bgh01274   B12     A6       6hpi     1    32
8             bgh01277   B12     A6       6hpi     1     9
9             bgh06140   B12     A6       6hpi     1    47
10            bgh05774   B12     A6       6hpi     1    42
# ... with 310,886 more rows
```

The key steps are to `gather()` the data into key-value pairs and 
then `separate()` the information from the key column into new columns.

```{r problem_1}
# add tidying code here
library(tidyverse)
raw_data %>% gather(key,value,-gene.name) %>% 
  separate(key, into = c('plant', 'fungus', 'time.point', 'rep'), sep = '_') -> 
  cleaned_data
cleaned_data

# report the cleaned_data by just naming it, uncomment the following line:
# cleaned_data
```

## Problem 2

You need to translate what you learned on the command line into the R world.

Which plant has the highest expression of any gene in the 6hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "6hpi"' | sort -k6nr | head -n 1 | cut -f2
```

```{r}
# translate to dplyr code

filter(cleaned_data, time.point=="6hpi") %>% 
  arrange(desc(value)) %>% 
  select(plant) %>%  
  slice(1)

```

Which plant / fungus pair has the highest expression in the 18hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "18hpi"' | sort -k6nr | head -n 1 | cut -f2,3
```

```{r}
# translate to dplyr code

filter(cleaned_data, time.point=="18hpi") %>%
  arrange(desc(value)) %>% 
  select(plant, fungus) %>% 
  select(1)
  # previous code: filter(row_number()==1)

```

## Problem 3

Identify the top 3 most consistently differentially expressed genes between the earliest and 
latest time points for each combination of `plant` and `fungus` strains.

- "Differential expression"" is the difference between `value` 
(i.e., gene expression level) between time points.

- "Consistency" is the smallest variance in `value` between replicates.

### Strategy

1. Create a new table from the cleaned data by moving each `hpi` value to a new column name, 
with counts for each in the column (hint: use a `tidyr` verb). It is helpful to reformat 
the `hpi` values by converting from e.g. `6hpi` to `hpi.6`. You can use `mutate` to do this, i.e.:

```{r reformat_time, eval=FALSE}
library(stringr)
# Version_1
cleaned_data %>% 
  mutate(time.point = str_replace(time.point, 'hpi', ''), 
         time.point = str_c('hpi.', time.point))-> cleaned_data_hpi
cleaned_data_hpi
```

2. Create a new column containing the expression difference between the relevant time points.

```{r new_column}
library(stringr)
# Version_1
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), 
                        time.point = str_c('hpi.', time.point))-> cleaned_data_hpi
cleaned_data_hpi %>% spread(time.point,value) %>%
  mutate(expr.diff = abs(hpi.24 - hpi.6), expr.diff.var = var(expr.diff)) -> expr_data
expr_data
```

3. Calculate summary statistics (mean and variance) of the expression differences 
by grouping (hint) the `gene.name`, and `plant` columns.


```{r substraction}
# Using substraction to determine the differentially expressed genes
expr_data %>% group_by(plant, fungus, gene.name) %>%
  summarise(expr.diff.mean = mean(expr.diff), 
            expr.diff.var=var(expr.diff)) -> expr_data_mean 
expr_data_mean 
```

4. Sort by these statistics and use the `dplyr` verb `slice` to pull the ones you want (i.e., the top 3). 
Note you will have to remove `gene.name` from the grouping so that sorting works.


```{r}
expr_data_mean %>% arrange(expr.diff.var) %>% 
 arrange(desc(expr.diff.mean)) %>%
 group_by(plant, fungus) %>%
 slice(1:3) -> top3_genes0
top3_genes0
```

```{r}
expr_data_mean %>% arrange(expr.diff.var) %>% 
 arrange(desc(expr.diff.mean)) %>%
 group_by(plant, fungus) %>%
 slice(1:2) -> top3_genes
```

## Problem 4

Now examine the above final data frame above and write a few sentences 
putting inline code in least 3 places. 
For example, There are `r nrow(mtcars)` rows of data in `mtcars`.


```{r, echo = FALSE}

filter(top3_genes0, grepl("B12", plant)) %>%  filter(., grepl("A6", fungus)) %>%
  ungroup() %>% 
  select(expr.diff.mean) %>% 
  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_B12_A6

filter(top3_genes0, grepl("pps", plant)) %>%  filter(., grepl("A6", fungus)) %>% 
  ungroup() %>% 
  select(expr.diff.mean) %>% 
  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_pps_A6

filter(top3_genes0, grepl("B12", plant)) %>%  filter(., grepl("K1", fungus)) %>%  
  ungroup() %>% 
  select(expr.diff.mean) %>%
  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_B12_K1

filter(top3_genes0, grepl("pps", plant)) %>%  filter(., grepl("K1", fungus)) %>% 
  ungroup() %>% 
  select(expr.diff.mean) %>%
  slice(1) %>% format(round(1), nsmall = 1) -> top_gene_pps_K1

# top3_genes_for_B12_A6
top3_genes0 %>% filter(plant =="B12", fungus =="A6") %>% ungroup() %>% select(gene.name) %>%
 t()-> top3_genes_for_B12_A6

# Other way giving the same result using subset():
# top3_genes0 %>% subset(., plant =="B12" & fungus =="A6", select=c(gene.name)) %>%
  #t()-> top3_genes_for_B12_A6
  
# top3_genes_for_B12_K1
top3_genes0 %>% filter(plant =="B12", fungus =="K1") %>% ungroup() %>% select(gene.name) %>% 
 t()-> top3_genes_for_B12_K1

# top3_genes_for_pps_A6
top3_genes0 %>% filter(plant =="pps", fungus =="A6") %>% ungroup() %>% select(gene.name) %>%
  t() -> top3_genes_for_pps_A6
  
#top3_genes_for_pps_K1
top3_genes0 %>% filter(plant =="pps", fungus =="K1") %>% ungroup() %>% select(gene.name) %>%
  t()-> top3_genes_for_pps_K1

```

My tidy dataset has `r ncol(top3_genes)` columns and `r nrow(top3_genes)` rows. 
The name of the columns are `r colnames(top3_genes)`.

The highest difference in expression of a gene in the B12 plant expressing 
the immune receptor MLA1-HA treated with effectors different
from the AVRA1 effector is: `r top_gene_B12_A6`.

The highest difference im expression of a gene in the pps plant 
(partially immunocompromised) treated with effectors different 
from the AVRA1 effector is: `r top_gene_pps_A6`.

The highest difference in expression of a gene in the B12 plant expressing 
the immune receptor MLA1-HA treated 
with the AVRA1 effector is: `r top_gene_B12_K1`.

The highest difference in expression of a gene in the pps plant 
(partially immunocompromised) treated with effectors different 
from the AVRA1 effector is: `r top_gene_pps_K1`.

The top 3 genes for the pair (B12 plant + A6 fungus) are: `r top3_genes_for_B12_A6`.

The top 3 genes for the pair (B12 plant + K1 fungus) are: `r top3_genes_for_B12_K1`.

The top 3 genes for the pair (B12 plant + A6 fungus) are: `r top3_genes_for_pps_A6`.

The top 3 genes for the pair (B12 plant + K1 fungus) are: `r top3_genes_for_pps_K1`.


## Problem 5

Plot the expression (`value`) by time (`hpi`) for the above genes. 
Format the plot as follows:


```{r preliminary_for_the_graphs}
# 3 top genes per pair plant/fungus, 4 pairs, triplicates, 
# 4 time points = 3 x 4 x 3 x 4 = 144
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', '')) %>%
  select(gene.name, plant, fungus, rep, time.point, value) %>%
  group_by(plant,fungus,gene.name)-> data_selected
filter(data_selected, gene.name %in% top3_genes$gene.name) -> all_data_for_graphs
all_data_for_graphs
```


1. Plot each point.
```{r question_5_1}
# Plot each point using geom_point (and also change the label of the x axis)
library(ggplot2)
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value)) + 
  geom_point() + xlab("time")

```
2. Connect the points with a smooth line 
```{r question_5_2}
# Connect the points with a smooth line using geom_smooth()
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, group = gene.name)) +
 geom_point()  + xlab("time") + geom_smooth(method = 'lm', se = F)
```

3. Create a facet for each `plant` / `bgh` pair 
```{r question_5_3}
# Create a facet for each plant/fungus pair using facet_wrap()
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), y = value, group = gene.name)) + 
  geom_point() + xlab("time") + facet_wrap(~plant~fungus) + geom_smooth(method = 'lm', se = F)
```
4. Assign a different color to each gene 
```{r question_5_4}
# Assign a color for each gene with color in aes
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), 
                                y = value, color = gene.name, group = gene.name)) + 
  geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + 
  geom_smooth(method = 'lm', se = F)
```

5. Assign a different shape to each replicate 

```{r question_5_5}
# Assigning a different shape to each replicate with shape in aes and scale_shape_manual()
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), 
                                y = value, color = gene.name, group = gene.name, shape=factor(rep))) + 
  geom_point()  + xlab("time") + facet_wrap(~plant~fungus) +  
  scale_shape_manual(values=c(21,24,8)) +
  geom_smooth(method = 'lm', se = F)
```

6. Add a descriptive title to the plot 

```{r question_5_6}
# Adding a title with ggtitle()
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), 
                                y = value, color = gene.name, 
                                group = gene.name, shape=factor(rep))) + 
  geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + 
  scale_shape_manual(values=c(21,24,8)) + 
  geom_smooth(method = 'lm', se = F) +
  ggtitle('Gene expression profile of the top 3 differentially 
          expressed genes per pair plant/fungus over the time course')
```

7. Use the "black & white" theme

```{r question_5_7}
# Theme black and white using theme_bw() (and also theme cowplot, theme_cowplot())

library(cowplot)
all_data_for_graphs$time.point <- as.numeric(all_data_for_graphs$time.point)
ggplot(all_data_for_graphs, aes(x = factor(time.point), 
                                y = value, color = gene.name, 
                                group = gene.name, shape=factor(rep))) + 
  geom_point()  + xlab("time") + facet_wrap(~plant~fungus) + 
  scale_shape_manual(values=c(21,24,8)) + 
  geom_smooth(method = 'lm', se = F) +
  ggtitle('Gene expression profile of the top 3 differentially 
          expressed genes per pair plant/fungus over the time course') + 
  theme_bw() + theme_cowplot()
```

Extra credit: add error bars to the plot (use `geom_errorbar`).

```{r 5.8 extra credit - variant 1}
# Since I was not sure which way was preferred for the representation of the error bars,
# I proposed 2 variants (I suspect that there may many other better variants)

# Variant 1 using geom_line() with geom_errorbar

all_data_for_graphs %>% group_by(gene.name,plant,fungus,time.point) %>% 
  summarise(expr.diff.mean=mean(value), standard.deviation = sd(value), 
            ymin = expr.diff.mean - standard.deviation, ymax = expr.diff.mean + standard.deviation) %>% 
  arrange(gene.name) -> all_data_for_graphs_with_sd

all_data_for_graphs_with_sd$time.point <- as.numeric(all_data_for_graphs_with_sd$time.point)

ggplot(all_data_for_graphs_with_sd, aes(x = factor(time.point), 
                                        y = expr.diff.mean, color = gene.name, group = gene.name)) +
  geom_line()  + xlab("time") +
  facet_wrap(~plant~fungus) +
   ggtitle('Gene expression profile of the top 3 differentially 
           expressed genes per pair plant/fungus over the time course') + 
  theme_bw() + theme_cowplot() +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.1) 

```

```{r 5.8 extra credit - variant 2}
# Variant 2 using geom_smooth() with geom_errorbar

all_data_for_graphs %>% group_by(gene.name,plant,fungus,time.point) %>% 
  summarise(expr.diff.mean=mean(value), standard.deviation = sd(value), 
            ymin = expr.diff.mean - standard.deviation, ymax = expr.diff.mean + standard.deviation) %>% 
  arrange(gene.name) -> all_data_for_graphs_with_sd

all_data_for_graphs_with_sd$time.point <- as.numeric(all_data_for_graphs_with_sd$time.point)

ggplot(all_data_for_graphs_with_sd, aes(x = factor(time.point), 
                                        y = expr.diff.mean, color = gene.name, group = gene.name)) +
  geom_point()  + xlab("time") + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~plant~fungus) +
   ggtitle('Gene expression profile of the top 3 differentially 
           expressed genes per pair plant/fungus over the time course') + 
  theme_bw() + theme_cowplot() +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.1)
```
